Modeling Interventions

### 2021-03-10 paper/paper.Rmd
finished a round of edits: 3133 words

keywords          : "early language learning; language acquisition; vocabulary development"
wordcount         : "3463"

@@

The core of the standard view of language learning is that it is a process of accumulation – individual experiences with words lead to their eventual acquisition. 
The more of these experiences a child receives, the faster their vocabulary grows. 
However, both children and words vary: children may learn faster or slower, and words can be more or less difficult to learn. 
Combining these ideas, the basic hypothesis is that a child’s vocabulary at a particular time should be predicted by their cumulative language exposure and learning rate, combined with the breadth and individual difficulties of the words that they have been exposed to.

The key contribution of such a baseline theory is that it establishes the link between the observable rates of words that children hear and their eventual knowledge of those words. 
For example, in principle the model could establish how many occurrences of the word “dog” are needed by the average child in order to start understanding (or producing) “dog.” 
Moreover, given a child’s vocabulary at a certain age, the model would provide an estimate of the child’s language ability.

This model outlines an approach that is very similar to Item-Response Theory [IRT; @embretson2013item], an influential and popular psychometric modeling framework for understanding the performance of test-takers (in our case language learners whose vocabulary is being assessed) as they are assessed with a particular set of test items (the particular words we ask about). 
In particular, IRT models jointly infer both how hard specific test items are and how accurate test takers are, recovering these latent parameters from the observed data. 
In the basic Rasch (or 1-parameter logistic) IRT model, a person i responds correctly to item j with probability determined by their ability ($\theta_i$) and the difficulty of item j ($d_j$):

@@

These parameters can easily be mapped onto the kind of accumulator model we have been describing: items are words (e.g. $d_j$ is the difficulty of word j), and $\theta_i$ is child $i$’s estimated latent language ability. 
Although in a typical IRT model both item difficulties and person abilities are standardized (assumed to be normally-distributed and centered on 0) and unit-free, in principle it is possible to map these scores to real-world distributions – as we would like to do – incorporating measured word frequencies and rates of children’s experienced input. 
With this mapping into absolute units, this kind of model can provide a quantitative linking hypothesis between measurements of input and learning. 
Recognizing the parallels between the standard model desiderata and IRT theory, we can use standard extensions of the 1PL (Rasch) IRT model to include both person-level and item-level covariates. Item-level (word-level) covariates of central interest in the standard model are estimates of word frequency in child-directed speech as well as lexical class [whether they are a noun, a predicate, or function word; @goodman2008does; @braginsky2019]. 
Person-level covariates of interest in a standard model might include sex, socioeconomic status variables, or other demographic information.

This model as described above formalizes an important and testable claim: is the amount of language experience the important predictor of learning for a child, or are there other age-related factors beyond experience? 
Put simply, is a 2-year-old different than a 1-year-old, aside from having twice as much experience with each word? 
This theoretical question can be answered by simple model comparison between two versions of the standard model, one in which age is included as a person-level covariate, and one in which age is instead included as a scalar on the prevalence of each item (i.e., in which the frequency of “dog” for a 2-year-old is twice that of “dog” for a 1-year-old). 
Using the IRT framework to formalize theoretical questions and assess them via model comparison is a strength of this approach. 
Below we report the results of this model comparison, fit to parent-reported vocabulary data from 5,492 English-speaking children in Wordbank. 
Each model estimated parameters for each child’s ability, and for the 680 words on the CDI Words and Sentences form. 
As covariates we included the lexical class of each word (function words, adjective, noun, verb, or other), and the estimated number of monthly tokens per word a child is likely to hear (range: 0.2 - 19,286 tokens), based on estimates from the CHILDES corpus of child-directed speech [@macwhinney2000childes]. 
An average child in this model (with input rate based on reported empirical means) is expected to receive 1,200 words per hour, 12 hours each day, for a total of 438,300 tokens per month, of which 285,200 tokens (65.1%) are accounted for by tokens of CDI words.

In Model 1, we included age as a person-level covariate, allowing for interactions of age (in months), lexical class, and monthly tokens per word. 
In Model 2, instead of including the child’s age as a person-level covariate, we scaled the expected tokens per month for each word by the child’s age, yielding an estimate of their lifetime tokens per word, and also allowed for interactions with lexical class. 
Model selection metrics (AIC and BIC) preferred the more complex Model 1, with a person-level age covariate separate from the item prevalence covariate. 
Predicted acquisition curves from Model 1 are shown in Figure 3, broken down by age, lexical class, and prevalence (left), showing for example that nouns are learned more rapidly than verbs or function words, and that there are interactions of prevalence and word type (e.g., more frequent nouns are learned earlier, but there is little effect of prevalence for verbs and function words). 
Moreover, the model predicts item-level acquisition curves (Figure 3, right), for example predicting that “ball” is learned earlier than “dog”, and that “go” is easier than “have.” 
Model 1 represents a new baseline standard model, which might be extended to include person-level groups such as SES and sex, as well as item-level covariates such as word length.

Any model in this framework also estimates the latent language ability of each child in the dataset -- a value correlated with age, and with a child’s total vocabulary size as estimated by the CDI, but that may be better correlated with the child’s quantity/quality of input received, or with their processing speed: relations that will require new datasets with concurrent measures of these variables in order to distinguish. 
For one lesson we have learned in synthesizing the standard model framework is that based on current datasets, some of the theories are currently indistinguishable from one another, the models unidentifiable. 
Thus, we conclude by briefly discussing the next steps needed to test and build upon the standard model.

@@

The goal of a computational theory like the "standard model" described here is to describe the quantitative predictions that arise from a set of assumptions. 
Often, however, it is the *failure* of such a model to predict observed patterns of data that is most useful, as these failures point the way forward towards future refinements [@vanrooij2020theory]. 
We discuss three potential failures here as signposts for ways forward.

**Leveraged learning and the role of processing**. As the model results above show, older children accumulate language from their input faster than younger children. 
Why? One potential reason is that independent developmental or maturational factors simply make older children better at remembering their experiences. 
Another possibility -- not mutually exclusive -– is that children *leverage* their knowledge of language to learn faster [@mitchell2009]. 
This leverage idea is consistent with many proposed mechanisms for language learning. For example, if children reason about new vocabulary items by exclusion [@markman1988children], they could get better at learning new words just by having more potential items to exclude from consideration. 
Yet a third possibility –- again not mutually exclusive with the others -– is that children are getting better at using the words they know. 
Several recent studies have found that a surprising proportion of variance in the rate of young children’s vocabulary growth is accounted for by the speed with which individual children are able to process familiar words [e.g., @fernald2006picking; @marchman2016early].
Disentangling these distinct explanations will require careful measurement of both vocabulary and the potential mechanisms of leverage [cf. @lewis2020role].

**Beyond vocabulary**. The "standard model" proposed here is a model of the accumulation of words, but language is a rich, complex system in which words are inflected morphologically and composed syntactically to express compositional meanings. 
In early conceptions of language learning, syntax and morphology were thought of as distinct and unconnected, but this conception has not been borne out empirically. 
Instead, evidence from large-scale data shows again and again that the language system is "tightly woven," with extremely tight correlations between the acquisition of words, morphology, and syntax [@wordbankbook]. 
Theoretically, accumulator models like the standard model are generic models of skill acquisition. 
Given the perspective that language learning is a form of skill acquisition [@chater2018language], such connections could lead the way towards extensions of the standard model to the accumulation of broader units of language like constructions.

**A theory for understanding acquisition in diverse contexts**. The "standard model" formalizes an implict assumption of the literature: namely, that early language learning follows the same general model for children in radically different circumstances. 
Yet this assumption could very well be false. 
Some children might learn more from overheard speech and others might learn more from child-directed speech [@sperry2019]. 
The kind of data necessary to test this assumption directly are only now being collected, for example in studies that rigorously track learning outcomes and amount of language input in diverse populations [e.g., comparisons between children growing up in low-income, rural, indigenous communities and those in higher-income Western contexts; @casillas2020].

>>>>

keywords          : "early language learning; language acquisition; vocabulary development"
wordcount         : "3133"

@@

The core of the standard view of language learning is that it is a process of accumulation – individual experiences with words lead to their eventual acquisition. 
The more of these experiences a child receives, the faster their vocabulary grows. 
However, both children and words vary: children may learn faster or slower, and words can be more or less difficult to learn. 
Combining these ideas, the basic hypothesis is that a child’s vocabulary at a particular time should be predicted by their cumulative language exposure and learning rate, combined with the breadth and individual difficulties of the words to which have been exposed.

The key contribution of such a baseline theory is that it establishes the link between the observable rates of words that children hear and their eventual knowledge of those words. 
For example, in principle, the model could establish how many occurrences of the word “dog” are needed by the average child in order to start understanding (or producing) “dog.” 
Moreover, given a child’s vocabulary at a certain age, the model would provide an estimate of the child’s language ability.

This model outlines an approach similar to Item-Response Theory [IRT; @embretson2013item], an influential and popular psychometric modeling framework for understanding the performance of test-takers (in our case language learners) as they are assessed with a particular set of test items (particular words). 
In particular, IRT models jointly infer both how hard specific test items are and how accurate test takers are, recovering these latent parameters from the observed data. 
In the basic Rasch (or 1-parameter logistic) IRT model, a person i responds correctly to item j with probability determined by their ability ($\theta_i$) and the difficulty of item j ($d_j$):

@@

These parameters can easily be mapped onto the kind of accumulator model we have been describing: items are words (e.g. $d_j$ is the difficulty of word j), and $\theta_i$ is child $i$’s estimated latent language ability. 
In typical IRT models both item difficulties and person abilities are standardized (assumed to be normally-distributed and centered on 0) and unit-free. 
However, in principle, it is possible to map these scores to real-world distributions incorporating measured word frequencies and rates of children’s experienced input. 
With this mapping into absolute units, this kind of model can provide a quantitative linking hypothesis between measurements of input and learning. 
Recognizing the parallels between the standard model desiderata and IRT theory, we can use standard extensions of the 1PL (Rasch) IRT model to include both person-level and item-level covariates. 
Item-level (word-level) covariates of central interest are estimates of word frequency in child-directed speech, as well as lexical class [e.g., noun, predicate, or function word; @goodman2008does; @braginsky2019]. 
Person-level covariates of interest might include sex, socioeconomic status variables, or other demographic information.

This model as described above formalizes an important and testable claim: is the amount of language experience the important predictor of learning for a child, or are there other age-related factors beyond experience? 
Put simply, is a 2-year-old different than a 1-year-old, aside from having twice as much experience with each word? 
This theoretical question can be answered by simple model comparison between two versions of the standard model, one in which age is included as a person-level covariate, and one in which age is included as a scalar on the prevalence of each item (i.e., in which the frequency of “dog” for a 2-year-old is twice that of “dog” for a 1-year-old). 
Using the IRT framework to formalize theoretical questions and assess them via model comparison is a strength of this approach.

Below we report the results of this model comparison, fit to parent-reported vocabulary data from 5,492 English-speaking children in Wordbank. 
Each model estimated parameters for each child’s ability, and for the 680 words on the CDI Words and Sentences form. 
Covariates were the lexical class of each word (function word, adjective, noun, verb, or other), and the estimated number of monthly tokens per word a child hears (range: 0.2 - 19,286 tokens), based on estimates from the CHILDES corpus of child-directed speech [@macwhinney2000childes]. 
This model expects an average child to receive 1,200 words per hour, 12 hours each day, for a total of 438,300 tokens per month, of which 285,200 tokens (65.1%) are accounted for by tokens of CDI words.

In Model 1, we included age as a person-level covariate, allowing for interactions of age (in months), lexical class, and monthly tokens per word. 
Model 2 dropped the age covariate and instead scaled the expected tokens per month for each word by child age, yielding an estimate of lifetime tokens per word, and also allowed for interactions with lexical class. 
Model selection metrics (AIC and BIC) preferred the more complex Model 1, with a person-level age covariate separate from the item prevalence covariate. 
Predicted acquisition curves from Model 1 are shown in Figure 3, broken down by age, lexical class, and prevalence (left), revealing that nouns are learned more rapidly than verbs or function words, and that there are interactions of prevalence and word type (e.g., more frequent nouns are learned earlier, but there is little effect of prevalence for verbs and function words). 
Moreover, the model predicts item-level acquisition curves (Figure 3, right): e.g., “ball” is learned earlier than “dog”, and “go” is easier than “have.” 
Model 1 represents a new baseline standard model, which might be extended to include person-level groups such as SES and sex, as well as item-level covariates such as word length.

These models also estimate a latent language ability for each child in the dataset. 
However, this ability is no doubt correlated with age, sex (females have larger vocabularies at young ages), and likely with features related to input quality. 
Estimating the contributions of these factors and determining whether/how language ability changes over developmental time is not possible with current datasets.
Thus, we conclude by discussing the next steps needed to test and build upon the standard model.

@@

The goal of a computational theory like the "standard model" described here is to describe the quantitative predictions that arise from a set of assumptions. 
Often, however, it is the *failure* of such a model to predict observed patterns of data that is most useful, as these failures point the way forward towards future refinements [@vanrooij2020theory]. 
We discuss three potential failures here.

**Leveraged learning and the role of processing**. As the model results above show, older children accumulate language from their input faster than younger children. 
There are at least three non-mutually exclusive potential reasons for this.
First, older children may be better at remembering their experiences due to independent developmental/maturational factors. 
Another possibility is that children *leverage* their knowledge of language to learn faster through any of several proposed mechanisms [@mitchell2009]. 
For example, if children reason about new words by exclusion [@markman1988children], they get better at learning new words just by having more potential items to exclude from consideration. 
A third possibility is that children are getting better at using the words they know. 
Several recent studies found that a surprising proportion of variance in the rate of young children’s vocabulary growth is accounted for by the speed with which individual children are able to process familiar words [e.g., @fernald2006picking; @marchman2016early].
Disentangling these distinct explanations will require careful measurement of both vocabulary and the potential mechanisms of leverage [cf. @lewis2020role].

**Beyond vocabulary**. The "standard model" proposed here models the accumulation of words.
Yet, of course, language is a rich, complex system in which words are inflected morphologically and composed syntactically to express compositional meanings. 
In early conceptions of language learning, syntax and morphology were thought of as distinct and unconnected, but this conception has not been borne out empirically. 
Instead, evidence from large-scale data shows again and again that the language system is "tightly woven," with extremely tight correlations between the acquisition of words, morphology, and syntax [@wordbankbook]. 
Theoretically, accumulator models like the standard model are generic models of skill acquisition. 
Given the perspective that language learning is a form of skill acquisition [@chater2018language], such connections could lead the way towards extensions of the standard model to the accumulation of broader units of language like constructions.

**A theory for understanding acquisition in diverse contexts**. The "standard model" formalizes an implict assumption of the literature: namely, that early language learning follows the same general model for children in radically different circumstances. 
Yet this assumption could very well be false. 
Some children might learn more from overheard speech and others might learn more from child-directed speech [@sperry2019]. 
The kind of data necessary to test this assumption directly are only now being collected, for example, in studies that rigorously track learning outcomes and amount of language input in diverse populations [e.g., comparisons between children growing up in low-income, rural, indigenous communities and those in higher-income Western contexts; @casillas2020].
